---
title: "Lab 3"
subtitle: Duke STA 240 Fall 2025
author: replace this text with your name
format: 
  pdf:
    toc: false
editor: 
  mode: source
---

In what follows, you may find some notation useful. Let $D$ denote your true disease status, and $T$ denote the result of your test. Then

$$
\begin{aligned}
    p       &= P(D=+)        \\
    f_-     &= P(T=-| D=+)   \\
    f_+     &= P(T=+| D=-)   \\
    1 - f_- &= P(T=+| D = +) \\
    1 - f_+ &= P(T=-| D = -).
\end{aligned}
$$


{{< pagebreak >}}

# Task 1

You can write text here.

```{r}
#| label: task-1

# You can add code here
```

You can type math here:

$$
P(A|B)=\frac{P(B|A)P(A)}{P(B)}.
$$

{{< pagebreak >}}

# Task 2

You can write text here.

```{r}
#| label: task-2

# You can add code here
```

You can type math here:

$$
P(A|B)=\frac{P(B|A)P(A)}{P(B)}.
$$

{{< pagebreak >}}

# Task 3

You can write text here.

```{r}
#| label: task-3

# You can add code here
```

You can type math here:

$$
P(A|B)=\frac{P(B|A)P(A)}{P(B)}.
$$

{{< pagebreak >}}

# Task 4

You can write text here.

```{r}
#| label: task-4

# You can add code here
```

You can type math here:

$$
P(A|B)=\frac{P(B|A)P(A)}{P(B)}.
$$

{{< pagebreak >}}

# Task 5

You can write text here.

```{r}
#| label: task-5

# You can add code here
```

You can type math here:

$$
P(A|B)=\frac{P(B|A)P(A)}{P(B)}.
$$
