---
title: "Lab 2"
subtitle: Duke STA 240 Fall 2025
author: replace this text with your name
format: 
  pdf:
    toc: false
editor: 
  mode: source
---


# Task 1

- **Scenario**: roll three *fair* dice;
- **Event**: all three dice show different numbers.

This code simulates 7,500 trials and computes the proportion of the time the event occurs:

```{r}
set.seed(37367)
nreps <- 7500
counter <- 0

for(i in 1:nreps){
  rolls <- sample(1:6, 3, replace = TRUE)
  
  if( length(unique(rolls)) == 3 ){
    counter <- counter + 1
  }
  
}

counter / nreps # proportion
```

In order to do the math, we apply $P(A) = \#(A) / \#(S)$ directly. The sample space $S$ is the set of all three-dice rolls. Thought of as three experiments in the sense of the counting principle, the number of ways we could mix-and-match the rolls is $\#(S) = 6 \times 6 \times 6 = 6^3 = 216$. The target event $A$ is the set of all three-dice rolls where all numbers are different (ie sampling without replacement), and so $\#(A) = 6 \times 5 \times 4 = 120$. Putting it into `R` gives:

```{r}
120 / 216
```

We conclude then that

$$
P(A) = \frac{6 \times 5 \times 4}{6^3} = \frac{120}{216} \approx 0.555,
$$

which agrees with our simulation.

{{< pagebreak >}}

# Task 2

You can write text here.

```{r}
#| label: task-2-sim

# You can add code here
```



```{r}
#| label: task-2-calc

# You can add code here
```

You can type math here:

$$
P(A) = 1 - P(A^c) = 1 - \frac{ \binom{48}{5} }{ \binom{52}{5} } \approx 0.34.
$$

{{< pagebreak >}}

# Task 3

You can write text here.

```{r}
#| label: task-3-sim

# You can add code here
```



```{r}
#| label: task-3-calc

# You can add code here
```

You can type math here:

$$
P(A) = 1 - P(A^c) = 1 - \frac{ \binom{48}{5} }{ \binom{52}{5} } \approx 0.34.
$$
